Diffeomorphic Directly Manipulated Free-Form Deformation Image Registration via Vector Field Flows

### 2012-01-05 Posters/NNS2011/myposter.tex
New Safety Suit album---good stuff.

\begin{center}
{\bf Multivariate analysis using SCCA}\\
\end{center}
\vspace{-1mm}
Sparse canonical correlation analysis (SCCA) adapts classical CCA to
situations where the number of predictors is much greater than the number
of subjects for inference of correlative relationships between different
``views'' of some underlying phenomenon. For gray matter density ($\mathbf{G}$) and
DTI-derived FA  ($\mathbf{F}$) views, SCCA solves the following
criterion \cite{Avants2010b}:
\vspace{-7mm}
\begin{center}
\begin{align*}
\underset{\omega_{\mathbf{G}}, \omega_{\mathbf{F}}}{\operatorname{argmax}} \left\{\omega_{\mathbf{G}} \mathbf{G}^\mathrm{T}\mathbf{F}\omega_{\mathbf{F}} - \lambda_{\mathbf{G}}\sum_{v_{\mathbf{G}}} |\omega_{\mathbf{G}}|_1 - \lambda_{\mathbf{F}}\sum_{v_{\mathbf{F}}} |\omega_{\mathbf{F}}|_1\right\}: \|\omega_{\mathbf{G}}\|, \|\omega_{\mathbf{F}}\| \leq 1.
\end{align*}
\end{center}

@@

\begin{center}
\begin{tabular}{p{11cm}p{11cm}}
{\makebox[11cm][c]{\bf Individual subject processing}\vspace{-3mm} } & 
{\makebox[11cm][c]{\bf Template generation}} \\
{\makebox[11cm][c]{\includegraphics[height=60mm]{individualProcessing.pdf}}}\vspace{3mm} &
{\makebox[11cm][c]{\includegraphics[height=60mm]{templateBuilding.pdf}}} \\
\small Processing of each anatomical (T1-weighted) image consists of the
following steps: 
\begin{itemize}
  \item bias correction,
  \item skull stripping using Atropos coupled with anatomical priors, and
  \item prior-based csf, white matter, and gray matter segmentation.  The
  latter yields the gray matter density image for each subject.
\end{itemize}
FA images are mapped to each anatomical image using the transform
obtained from registration of the average diffusion image to the 
corresponding T1. 
& 
\small Statistical analysis requires normalization to a standard space provided
by the total template.  All 17 controls and all 16 TBI patients were used
to generate their respective unbiased templates.  These two templates were then
used to generate the final/standard template.  Thus the mapping from the individual 
gray matter density (GMD) image to the total template is given by
\begin{align*}
  \mathrm{GMD}_{subject}\,\,\mathbf{Id}\,\, \mathrm{T1}_{subject}\leadsto \mathrm{template}_{population}\leadsto \mathrm{template}_{total}.
\end{align*}
Similarly, the mapping for each subject's FA image to the total template is given
by the 
transform composition
\begin{align*}
  \mathrm{FA}_{subject}\rightarrow\mathrm{T1}_{subject}\leadsto \mathrm{template}_{population}\leadsto \mathrm{template}_{total}.
\end{align*}
\\
\end{tabular}
\end{center}

@@

\begin{center}
\begin{tabular}{p{6cm}cc} \vspace{-70mm}
{\makebox[6cm][c]{\includegraphics[height=50mm]{scca.pdf} }} \
\small The significant correlations derived from the GMD/FA views
vs. diagnostic view provide significant regions for finding
correlations between GMD and FA. & {           } &
{\includegraphics[height=80mm]{results.pdf}}
\end{tabular}
\end{center}

@@

 SCCA demonstrates significant differences between the control and patient groups in both the FA ($p < 0.002$) and gray matter ($p < 0.04$)	that	are	widespread	but	largely	focus	on	thalamocortical networks related to the limbic system. Specific regional differences included the medial thalamic nuclei, hypothalamus, amygdala, hippocampus, anterior cingulate cortex, orbitofrontal cortex and fornix. Using these SCCA-identified regions, we demonstrate a strong correlation ($p < 0.08$) of the degree of injury in WM and GM within the patient group.

@@

  \begin{center}
  {\normalsize \bf Open source software} \\
{\vspace{3mm}}
  \begin{tabular}{cc}
  \includegraphics[height=12mm]{ants_logo.png} &
  \includegraphics[height=11mm]{itkLogo.jpg} \\
%  {Advanced Normalization Tools} & {Insight Toolkit} \\
  {\tiny http://www.picsl.upenn.edu/ANTs} &
%  {\tiny http://www.cs.ucl.ac.uk/research/medic/camino} &
  {\tiny http://www.itk.org/} \\
  \end{tabular}
  \end{center}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
  \headerbox{Acknowledgments}{name=acknowledgments,column=1,below=conclusions,above=bottom}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
  \vspace{1mm}
  \scriptsize
  \begin{itemize}
  \item Partial funding support from NZ Breacher study
  \item This project is also funded, in part, under a grant with the Pennsylvania 
  Department of Health. The Department specifically disclaims responsibility 
  for any analyses, interpretations, or conclusions.
  \end{itemize}
  }

>>>>

\begin{center}
{\bf Multivariate analysis using SCCA}\\
\end{center}
\vspace{-1mm}
Sparse canonical correlation analysis (SCCA) adapts classical CCA to
situations where the number of predictors is much greater than the number
of subjects for inference of correlative relationships between different
``views'' of some underlying phenomenon. For gray matter density ($\mathbf{G}$) and
DTI-derived  ($\mathbf{F}$) views, SCCA solves the following
criterion \cite{Avants2010b}:
\vspace{-7mm}
\begin{center}
\begin{align*}
\underset{\omega_{\mathbf{G}}, \omega_{\mathbf{F}}}{\operatorname{argmax}} \left\{\omega_{\mathbf{G}} \mathbf{G}^\mathrm{T}\mathbf{F}\omega_{\mathbf{F}} - \lambda_{\mathbf{G}}\sum_{v_{\mathbf{G}}} |\omega_{\mathbf{G}}|_1 - \lambda_{\mathbf{F}}\sum_{v_{\mathbf{F}}} |\omega_{\mathbf{F}}|_1\right\}: \|\omega_{\mathbf{G}}\|, \|\omega_{\mathbf{F}}\| \leq 1.
\end{align*}
\end{center}

@@

\begin{center}
\begin{tabular}{p{11cm}p{11cm}}
{\makebox[11cm][c]{\bf Individual subject processing}\vspace{-3mm} } & 
{\makebox[11cm][c]{\bf Template generation}} \\
{\makebox[11cm][c]{\includegraphics[height=60mm]{individualProcessing.pdf}}}\vspace{3mm} &
{\makebox[11cm][c]{\includegraphics[height=60mm]{templateBuilding.pdf}}} \\
\small Processing of each anatomical (T1-weighted) image consists of the
following steps: 
\begin{itemize}
  \item bias correction,
  \item skull stripping using Atropos coupled with anatomical priors, and
  \item prior-based cerebrospinal fluid, white and gray matter segmentation.  The
  latter yields the gray matter density (GMD) image for each subject.
\end{itemize}
DTI-derived fractional anisotropy (FA) images are mapped to each anatomical image using the transform
obtained from registration of the average diffusion image to the 
corresponding T1. 
& 
\small Statistical analysis requires normalization to a standard space provided
by the total template.  All 17 controls and all 16 TBI patients were used
to generate their respective unbiased templates.  These two templates were then
used to generate the final/standard template.  Thus the mapping from the individual 
GMD image to the total template is given by
\begin{align*}
  \mathrm{GMD}_{subject}\,\,\mathbf{Id}\,\, \mathrm{T1}_{subject}\leadsto \mathrm{template}_{population}\leadsto \mathrm{template}_{total}.
\end{align*}
Similarly, the mapping for each subject's FA image to the total template is given
by the 
transform composition
\begin{align*}
  \mathrm{FA}_{subject}\rightarrow\mathrm{T1}_{subject}\leadsto \mathrm{template}_{population}\leadsto \mathrm{template}_{total}.
\end{align*}
\\
\end{tabular}
\end{center}

@@

\begin{center}
\begin{tabular}{p{6cm}cc} \vspace{-70mm}
{\makebox[6cm][c]{\includegraphics[height=50mm]{scca.pdf} }} \
\small The significant correlations derived from the GMD/FA views
vs. diagnostic view provide significant regions for finding
correlations between GMD and FA. & {           } &
{\includegraphics[height=80mm]{resultsPartial.pdf}}
\end{tabular}
\end{center}

@@

 SCCA demonstrates significant differences between the control and patient groups in both the FA ($p < 0.002$) and gray matter ($p < 0.04$)	that	are	widespread 	but	largely	focus	on	 thalamocortical networks related to the limbic system. Specific regional differences included the medial thalamic nuclei, hypothalamus, amygdala, hippocampus, anterior cingulate cortex, orbitofrontal cortex and fornix. Using these SCCA-identified regions, we demonstrate a strong correlation ($p < 0.01$) of the degree of injury in WM and GM within the patient group.

@@

  \begin{center}
  {\small \bf Open source software} \\
{\vspace{3mm}}
  \begin{tabular}{cc}
  \includegraphics[height=12mm]{ants_logo.png} &
  \includegraphics[height=11mm]{itkLogo.jpg} \\
%  {Advanced Normalization Tools} & {Insight Toolkit} \\
  {\tiny http://www.picsl.upenn.edu/ANTs} &
%  {\tiny http://www.cs.ucl.ac.uk/research/medic/camino} &
  {\tiny http://www.itk.org/} \\
  \end{tabular}
  \end{center}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
  \headerbox{Acknowledgments}{name=acknowledgments,column=1,below=conclusions,above=bottom}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
  \vspace{1mm}
  \smaller
   Partial funding support was provided by US ARMY Medical Research
   and Materiel Command grant W81XWH-09-2-0055.
   This project is also funded, in part, under a grant with the Pennsylvania 
  Department of Health. The Pennsylvania Department of Health specifically disclaims responsibility 
  for any analyses, interpretations, or conclusions.
  }

### 2012-01-05 WBIR2012/paper.tex
ENH:  Added reference.
New Safety Suit album---good stuff.

\begin{abstract}
Motivated by previous work \cite{tustison2009} and 
recent diffeomorphic image registration developments in which the 
characteristic velocity field is represented by spatiotemporal B-splines 
\cite{de-craene2011}, we present a diffeomorphic B-spline-based
image registration algorithm combining and extending these techniques.  
The advancements of the proposed framework
over previous work include analytical integration of
the continuous B-spline velocity field, potential spatial weighting 
of the gradient permitting (among other things) enforcement of 
stationary boundary conditions, multiresolution in terms of the B-spline
mesh, facilitating the incorporation of landmark information, 
and the possibility of modeling temporal periodicity.  In addition to
theoretical and practical discussions of our contribution, we also describe 
its parallelized implementation as open source in the Insight Toolkit.  
\keywords{B-splines, DMFFD, diffeomorphism, ITK}
\end{abstract}

@@

Other important image registration research
reflected increased emphasis on topological transformation considerations
in modeling biological/physical systems where topology is 
consistent throughout the course of deformation or a 
homeomorphic relationship is assumed between image domains.
Methods such as LDDMM \cite{beg2005} optimize time-varying velocity field 
flows to yield diffeomorphic transformations.  Alternatively, the FFD 
variant reported in \cite{rueckert2006} enforced diffeomorphic transforms
by concatenating multiple FFD transforms, each of which is constrained
to describe a one-to-one mapping.
Recently, the work of 
\cite{de-craene2011} combined these registration concepts into a single
framework called {\em temporal free-form deformation} in which the 
time-varying velocity field characteristic of LDDMM-style algorithms
is modeled using a 4-D B-spline object (3-D + time).  Integration of 
the velocity field yields the mapping between parameterized time points.

In this work, we describe our extension to these methods.  Similar to 
\cite{de-craene2011}, we also use an $N$-D + time B-spline object to 
represent the characteristic velocity field.  However, we use the 
directly manipulated free-form deformation optimization formulation to improve 
convergence during the course of optimization.  This also facilitates
modeling temporal periodicity and
the enforcement of stationary boundaries consistent with diffeomorphic
transforms.
We also incorporate B-spline mesh multi-resolution capabilities
for increased control during registration progress.  Of note, whereas the 
previous work described in \cite{de-craene2011} discretized the velocity field before 
performing numerical integration (which reduces the advantages of the continuous
B-spline representation including the decrease in accuracy associated with
numerical integration and increased computational time) we show how the continuous B-spline 
representation permits analytical integration for increased accuracy.  
Most importantly, we also describe
our parallelized algorithmic implementation as open source available through the Insight Toolkit.%
\footnote{
http://www.itk.org/
}

@@

Briefly, as with other diffeomorphic formulations based on vector flows, we
assume the diffeomorphism, $\phi$, is defined on the image domain, $\Omega$, 
with stationary boundaries such that $\phi( \partial \Omega) = \mathbf{Id}$.
$\phi$ is generated as the solution of the ordinary differential
equation 
\begin{align}
  \label{eq:ode}
  \frac{d\phi(\mathbf{x}, t)}{dt} = v(\phi(\mathbf{x}, t), t),\,\,\phi( \mathbf{x}, 0 ) = \mathbf{x}
\end{align}
where $v$ is a (potentially) time-dependent smooth field, $v : \Omega \times t 
\rightarrow \mathrm{R}^d$ parameterized by $t \in [0,1]$.  Diffeomorphic mappings
between parameterized time points $\{t_a,t_b\} \in [0,1]$ 
are obtained from  Eq. (\ref{eq:ode}) through the following integration
\begin{align}
  \label{eq:integral}
\phi(\mathbf{x},t_b) &= \phi(\mathbf{x},t_a) + \int_{t_a}^{t_b} v(\phi(\mathbf{x}), t) dt.
\end{align}

In the case of 
$d$-dimensional registration, we can represent the time-dependent velocity field  
as a $(d + 1)$-dimensional B-spline object
\begin{align}
v(\phi(\mathbf{x},t), t) = \sum_{i_1=1}^{X_1}\ldots\sum_{i_d=1}^{X_d}\sum_{i_t=1}^T v_{i_1,\ldots,i_d,i_t} B_{i_t}(t) \prod_{j=1}^d B_{i_j}(x_j)
\end{align}
where $v_{i_1,\ldots,i_d,i_t}$ is a $(d+1)$-dimensional control point lattice
characterizing the velocity field and $B(\cdot)$ are the univariate B-spline
basis functions separately modulating the solution in each parametric dimension.
Combining Eqns. (\ref{eq:ode}) and (\ref{eq:integral}) we obtain
\begin{align}
\phi(\mathbf{x},t_b) = \phi(\mathbf{x},t_a) + \int_{t_a}^{t_b} \left( \sum_{i_1=1}^{X_1}\ldots\sum_{i_d=1}^{X_d}\sum_{i_t=1}^T 
    v_{i_1,\ldots,i_d,i_t} B_{i_t}(t)\prod_{j=1}^d B_{j}(x_j) \right) dt 
\end{align}
Moving the integral inside the summations
\begin{align}
\label{eq:velocityfield}
\phi(\mathbf{x},t_b) = \phi(\mathbf{x},t_a) + \sum_{i_1=1}^{X_1}\ldots\sum_{i_d=1}^{X_d} 
        \underbrace{\left( \sum_{i_t=1}^T v_{i_1,\ldots,i_d,i_t} \int_{t_a}^{t_b} B_{i_t}(t) dt \right)}_{\phi'_{i_1,\ldots,i_d}} \prod_{j=1}^d B_{i_j}(x_j) 
\end{align}
demonstrates the sole dependency on the B-spline basis functions in
the temporal parametric dimension.  The integral is easily calculated
as the B-spline basis functions are piecewise polynomials.
Also, it is seen that the 
quantity inside the parentheses comprises the $d$-dimensional 
control point lattice defining the diffeomorphic mapping
defined between time points $[t_a,t_b]$ obtained by 
integration.  Furthermore, supposing the initial diffeomorphism,
$\phi(\mathbf{x}, t_a)$, is similarly parameterized, i.e.
\begin{align}
  \phi(\mathbf{x}, t_a) = \sum_{i_1=1}^{X_1}\ldots\sum_{i_d=1}^{X_d} \phi^{t_a}_{i_1,\ldots,i_d}\prod_{j=1}^d B_{i_j}(x_j), 
\end{align}
the diffeomorphism at $t_b$ is obtained by simple addition of the two control point 
lattices,
\begin{align}
\phi(\mathbf{x},t_b) = \sum_{i_1=1}^{X_1}\ldots\sum_{i_d=1}^{X_d} \left(\phi^{t_a}+
\phi' \right)_{i_1,\ldots,i_d} \prod_{j=1}^d B_{i_j}(x_j).
\end{align}

@@

As mentioned previously, the registration algorithm has been implemented in the
Insight Toolkit and consists of the following major classes: 
\begin{itemize}
  \item \verb#itk::TimeVaryingBSplineVelocityFieldIntegrationImageFilter#
  \item \verb#itk::TimeVaryingBSplineVelocityFieldTransform#
  \item \verb#itk::TimeVaryingBSplineVelocityFieldImageRegistrationMethod#
  \item \verb#itk::TimeVaryingBSplineVelocityFieldTransformParametersAdaptor#
  \item \verb#itk::BSplineScatteredDataPointSetToImageFilter#
\end{itemize}
The \verb#TimeVaryingBSplineVelocityFieldIntegrationImageFilter# class implements
the integration described by Eqn. (\ref{eq:velocityfield}).  Given a 
velocity field control point lattice as 
input and the lower and upper integration limits, integration is performed 
in a multi-threaded fashion (since each control point in $\phi'_{i_1,\ldots,i_d}$
can be integrated separately).  The \verb#TimeVaryingBSplineVelocityFieldTransform# 
is derived from the base transform class which handles the mapping of geometric primitives
for warping images.  The DMFFD gradient calculation  is handled by the class \verb#BSplineScatteredDataPointSetToImageFilter# (cf. Eqn. \ref{eq:dmffd} and line 11
of Algorithm \ref{alg1}) which takes as input the dense similarity metric and an optional
weighting for each gradient sample.  This permits enforcement of stationary physical
boundaries by specifying a zero gradient on the boundaries and a large weighting.
Coordinating all the elements of image registration is the
\verb#TimeVaryingBSplineVelocityFieldImageRegistrationMethod# class which
encapsulates Algorithm \ref{alg1}.  The resolution scheduling is determined
by the \verb#TimeVaryingBSplineVelocityFieldTransformParametersAdaptor# class.
We provide access to the new ITK 
registration framework (including the B-spline velocity field transform) 
through the command line module \verb#hormigita# available both in ANTs%
\footnote{
http://www.picsl.upenn.edu/ANTs
}
and accompanied by a technical report offered through the Insight Journal
\cite{tustison2012}.%
\footnote{
http://www.insight-journal.org
}
The interested reader should consult \cite{tustison2012} for further details on 
actual usage.

@@

\section{Discussion and Conclusions}
This work constitutes an advantageous combination of the continuous aspects of
B-splines with the diffeomorphic registration framework via vector field 
flows.  In contrast to previous work, we take advantage of the continuous 
aspect of the velocity field representation by analytical integration for
a more accurate diffeomorphic transform derivation.  We also incorporate
DMFFD optimization of the B-spline velocity field which facilitates 
convergence and permits enforcement of stationary boundary conditions.  
While not discussed, further advantages include incorporation of temporal 
periodicity in dealing with the possibility of multiple images describing
periodic motion (e.g. cardiac or pulmonary motion).  Although not discussed
in this work, a future publication will explore these possibilities in 
greater detail.

>>>>

\begin{abstract}
Motivated by previous work \cite{tustison2009} and 
recent diffeomorphic image registration developments in which the 
characteristic velocity field is represented by spatiotemporal B-splines 
\cite{de-craene2011}, we present a diffeomorphic B-spline-based
image registration algorithm combining and extending these techniques.  
The advancements of the proposed framework
over previous work include a preconditioned gradient descent algorithm, 
potential spatial weighting 
of the gradient permitting (among other things) enforcement of 
stationary boundary conditions, multiresolution in terms of the B-spline
mesh, facilitating the incorporation of landmark information, 
and the possibility of modeling temporal periodicity.  In addition to
theoretical and practical discussions of our contribution, we also describe 
its parallelized implementation as open source in the Insight Toolkit.  
\keywords{B-splines, DMFFD, diffeomorphism, ITK}
\end{abstract}

@@

Other important image registration research
reflected increased emphasis on topological transformation considerations
in modeling biological/physical systems where topology is 
consistent throughout the course of deformation or a 
homeomorphic relationship is assumed between image domains.
Methods such as LDDMM \cite{beg2005} optimize time-varying velocity field 
flows to yield diffeomorphic transformations.  Alternatively, the FFD 
variant reported in \cite{rueckert2006} enforced diffeomorphic transforms
by concatenating multiple FFD transforms, each of which is constrained
to describe a one-to-one mapping.  Another FFD registration
incorporated the recent log-Euclidean framework for enforcing diffeomorphic
transformations \cite{Modat2011}.
Recently, the work of 
\cite{de-craene2011} combined these registration concepts into a single
framework called {\em temporal free-form deformation} in which the 
time-varying velocity field characteristic of LDDMM-style algorithms
is modeled using a 4-D B-spline object (3-D + time).  Integration of 
the velocity field yields the mapping between parameterized time points.

In this work, we describe our extension to these methods.  Similar to 
\cite{de-craene2011}, we also use an $N$-D + time B-spline object to 
represent the characteristic velocity field.  However, we use the 
directly manipulated free-form deformation optimization formulation to improve 
convergence during the course of optimization.  This also facilitates
modeling temporal periodicity and
the enforcement of stationary boundaries consistent with diffeomorphic
transforms.
We also incorporate B-spline mesh multi-resolution capabilities
for increased control during registration progress.  
Most importantly, we also describe
our parallelized algorithmic implementation as open source available through the Insight Toolkit.%
\footnote{
http://www.itk.org/
}

@@

Briefly, as with other diffeomorphic formulations based on vector flows, we
assume the diffeomorphism, $\phi$, is defined on the image domain, $\Omega$, 
with stationary boundaries such that $\phi( \partial \Omega) = \mathbf{Id}$.
$\phi$ is generated as the solution of the ordinary differential
equation 
\begin{align}
  \label{eq:ode}
  \frac{d\phi(\mathbf{x}, t)}{dt} = v(\phi(\mathbf{x}, t), t),\,\,\phi( \mathbf{x}, 0 ) = \mathbf{x}
\end{align}
where $v$ is a (potentially) time-dependent smooth field, $v : \Omega \times t 
\rightarrow \mathrm{R}^d$ parameterized by $t \in [0,1]$.  Diffeomorphic mappings
between parameterized time points $\{t_a,t_b\} \in [0,1]$ 
are obtained from  Eq. (\ref{eq:ode}) through integration of the transport
equation, viz.
\begin{align}
  \label{eq:integral}
\phi(\mathbf{x},t_b) &= \phi(\mathbf{x},t_a) + \int_{t_a}^{t_b} v(\phi(\mathbf{x}), t) dt.
\end{align}

In the case of 
$d$-dimensional registration, we can represent the time-dependent velocity field  
as a $(d + 1)$-dimensional B-spline object
\begin{align}
v(\mathbf{x}, t) = \sum_{i_1=1}^{X_1}\ldots\sum_{i_d=1}^{X_d}\sum_{i_t=1}^T v_{i_1,\ldots,i_d,i_t} B_{i_t}(t) \prod_{j=1}^d B_{i_j}(x_j)
\end{align}
where $v_{i_1,\ldots,i_d,i_t}$ is a $(d+1)$-dimensional control point lattice
characterizing the velocity field and $B(\cdot)$ are the univariate B-spline
basis functions separately modulating the solution in each parametric dimension.

Although various methods exist for solving Eqns. (\ref{eq:ode}) and (\ref{eq:integral}),
we use $4^{th}$-order Runge-Kutta, i.e.
\begin{align}
  \phi_{n+1} &= \phi_{n} + \frac{1}{6}\left( k_1 + 2k_2 + 2k_3 + k_4 \right) \\
  t_{n+1} &= t_{n} + \Delta t \\ 
  \phi_0 &= \phi(t_0) 
\end{align}
where
\begin{align}
  k_1 &= v\left( \phi_{n}, t_{n} \right)\Delta t \\
  k_2 &= v\left( \phi_{n} + \frac{k_1}{2}, t_{n} + \frac{\Delta t}{2} \right)\Delta t \\
  k_3 &= v\left( \phi_{n} + \frac{k_2}{2}, t_{n} + \frac{\Delta t}{2} \right)\Delta t \\
  k_4 &= v\left( \phi_{n} + k_3, t_{n} + \Delta t \right)\Delta t
\end{align}  
which provides a less computationally expensive, more stable alternative than other numerical
methods.

%Combining Eqns. (\ref{eq:ode}) and (\ref{eq:integral}) we obtain
%\begin{align}
%\phi(\mathbf{x},t_b) = \phi(\mathbf{x},t_a) + \int_{t_a}^{t_b} \left( \sum_{i_1=1}^{X_1}\ldots\sum_{i_d=1}^{X_d}\sum_{i_t=1}^T 
%    v_{i_1,\ldots,i_d,i_t} B_{i_t}(t)\prod_{j=1}^d B_{j}(x_j) \right) dt 
%\end{align}
%Moving the integral inside the summations
%\begin{align}
%\label{eq:velocityfield}
%\phi(\mathbf{x},t_b) = \phi(\mathbf{x},t_a) + \sum_{i_1=1}^{X_1}\ldots\sum_{i_d=1}^{X_d} 
%        \underbrace{\left( \sum_{i_t=1}^T v_{i_1,\ldots,i_d,i_t} \int_{t_a}^{t_b} B_{i_t}(t) dt \right)}_{\phi'_{i_1,\ldots,i_d}} \prod_{j=1}^d B_{i_j}(x_j) 
%\end{align}
%demonstrates the sole dependency on the B-spline basis functions in
%the temporal parametric dimension.  The integral is easily calculated
%as the B-spline basis functions are piecewise polynomials.
%Also, it is seen that the 
%quantity inside the parentheses comprises the $d$-dimensional 
%control point lattice defining the diffeomorphic mapping
%defined between time points $[t_a,t_b]$ obtained by 
%integration.  Furthermore, supposing the initial diffeomorphism,
%$\phi(\mathbf{x}, t_a)$, is similarly parameterized, i.e.
%\begin{align}
%  \phi(\mathbf{x}, t_a) = \sum_{i_1=1}^{X_1}\ldots\sum_{i_d=1}^{X_d} \phi^{t_a}_{i_1,\ldots,i_d}\prod_{j=1}^d B_{i_j}(x_j), 
%\end{align}
%the diffeomorphism at $t_b$ is obtained by simple addition of the two control point 
%lattices,
%\begin{align}
%\phi(\mathbf{x},t_b) = \sum_{i_1=1}^{X_1}\ldots\sum_{i_d=1}^{X_d} \left(\phi^{t_a}+
%\phi' \right)_{i_1,\ldots,i_d} \prod_{j=1}^d B_{i_j}(x_j).
%\end{align}

@@

As mentioned previously, the registration algorithm has been implemented in the
Insight Toolkit and consists of the following major classes: 
\begin{itemize}
  \item \verb#itk::TimeVaryingVelocityFieldIntegrationImageFilter#
  \item \verb#itk::TimeVaryingBSplineVelocityFieldTransform#
  \item \verb#itk::TimeVaryingBSplineVelocityFieldImageRegistrationMethod#
  \item \verb#itk::TimeVaryingBSplineVelocityFieldTransformParametersAdaptor#
  \item \verb#itk::BSplineScatteredDataPointSetToImageFilter#
\end{itemize}
The \verb#TimeVaryingVelocityFieldIntegrationImageFilter# class implements
the Runge-Kutta integration described earlier.  Given a 
velocity field control point lattice as 
input and the lower and upper integration limits, integration is performed 
in a multi-threaded fashion (since each point in the domain
can be integrated separately).  The \verb#TimeVaryingBSplineVelocityFieldTransform# 
is derived from the base transform class which handles the mapping of geometric primitives
for warping images.  The DMFFD gradient calculation  is handled by the class \verb#BSplineScatteredDataPointSetToImageFilter# (cf. Eqn. \ref{eq:dmffd} and line 11
of Algorithm \ref{alg1}) which takes as input the dense similarity metric and an optional
weighting for each gradient sample.  This permits enforcement of stationary physical
boundaries by specifying a zero gradient on the boundaries and a large weighting.
Coordinating all the elements of image registration is the
\verb#TimeVaryingBSplineVelocityFieldImageRegistrationMethod# class which
encapsulates Algorithm \ref{alg1}.  The resolution scheduling is determined
by the \verb#TimeVaryingBSplineVelocityFieldTransformParametersAdaptor# class.
We provide access to the new ITK 
registration framework (including the B-spline velocity field transform) 
through the command line module \verb#antsRegistration# available both in ANTs%
\footnote{
http://www.picsl.upenn.edu/ANTs
}
and accompanied by a technical report offered through the Insight Journal
\cite{tustison2012}.%
\footnote{
http://www.insight-journal.org
}
The interested reader should consult \cite{tustison2012} for further details on 
actual usage.

@@

\section{Discussion and Conclusions}
This work constitutes an advantageous combination of the continuous aspects of
B-splines with the diffeomorphic registration framework via vector field 
flows.  We incorporate
DMFFD optimization of the B-spline velocity field which facilitates 
convergence and permits enforcement of stationary boundary conditions.  
While not discussed, further advantages include incorporation of temporal 
periodicity in dealing with the possibility of multiple images describing
periodic motion (e.g. cardiac or pulmonary motion).  Although not discussed
in this work, a future publication will explore these possibilities in 
greater detail.
